---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# DgeaHeatmap

<!-- badges: start -->
<!-- badges: end -->

The goal of DgeaHeatmap is to enable R users to generate heatmaps more easily and to assist with preprocessing of read counts. Furthermore, the package is aimed at simplifying the extraction of raw read counts from .dcc and .pkc files generated by Nanostring GeoMx DSP.

## Installation

You can install the development version of DgeaHeatmap from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("leolanci/Dgea_Heatmap_Package")
```

## Usage
![Workflow.](SCheme_BioinformaticsPaper (1).png)
```{r ImageREADME, echo=FALSE, fig.cap="Step-by-step workflow of data processing using DgeaHeatmap"}
#print(knitr::include_graphics("C:/Users/llancell/Downloads/SCheme_BioinformaticsPaper (6).jpeg"))
#print("C:/Users/llancell/Downloads/SCheme_BioinformaticsPaper (6).jpeg")
#! [Workflow.](/Users/llancell/Downloads/SCheme_BioinformaticsPaper (6).jpeg)
#library(png)
#library(grid)
#img <- readPNG("C:/Users/llancell/Downloads/SCheme_BioinformaticsPaper (1).png")
# grid.raster(img)
```


### Building Heatmaps

**This is a basic example that shows you how to solve a common problem:**

```{r input normalizedReadCountsExample, echo=FALSE, results='hide'}
library(testthat)
input_data <- read.csv(test_path("1_Counts_All_Regions_All_Ages_Pos_Neg.csv"))
del_layer6 <- grep("layer", colnames(input_data))
input_data <- input_data[, -del_layer6]
```


```{r exampleBuildMatrix, echo=TRUE, results='hide'}
library(DgeaHeatmap)
x <- 1
matrixCounts <- build_matrix(input_data, x)

```
```{r showTable, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, out.height="60%", out.width="60%"}
knitr::kable(head(matrixCounts[, 1:4]),"pipe")
```
```{r exampleIndividualMatrix, echo=TRUE}
parameter1 = "cortex"
parameter2 = "pos"
factors_for_individual_matrix = list(parameter1, parameter2)
indiMatrix <- individual_matrix(factors_for_individual_matrix, matrixCounts)
```
```{r showIndividualMatrix, echo=FALSE}
knitr::kable(head(indiMatrix[, 1:9]),"pipe")
```

**Filtering the matrix for only an x amount of most variably expressed genes:**

```{r exampleFilteringForTopExprGenes, echo=TRUE}
top_number_of_genes <- 500
varGenesMatrix <- filtering_for_top_exprGenes(indiMatrix, top_number_of_genes)
print(nrow(varGenesMatrix))
```

**Using Z-score scaling to scale the values of the matrix:**

```{r exampleZScoreScaling, echo=TRUE, results='hide'}
scaled_counts <- scale_counts(varGenesMatrix)
```

**How to visualize the data distribution of the scaled counts:**

```{r exampleShowDataDistribution, echo=TRUE}
show_data_distribution(scaled_counts)
```
**Generating an elbow plot to choose k for k-mean generation:**

```{r exampleElbowPlot, echo=TRUE}
seed <- 1 # setting a seed for a reproducible outcome
elbow_plot(seed, scaled_counts)
```
As desired, the samples can further be summarized as biological replicates to help generate more clearly arranged heatmaps and to give a better overview.
```{r exampleSummariseBioReplicates, echo=TRUE, results='hide'}
probes <- list("P0_cortex_Iba1_pos", "P5_cortex_Iba1_pos", "P15_cortex_Iba1_pos")
sumBioRepsMatrix <- summarise_bio_replicates(scaled_counts, probes)
```
```{r showSummedBioReplicates, echo=FALSE}
knitr::kable(head(sumBioRepsMatrix[, 1:3]),"pipe")
```
**Generating K-means for clustering in the heatmap:**

```{r exampleKmeanGeneration, echo=TRUE}
seed <- 1 #set a seed for reproducibility, but try different seeds first
k_clusters <- 2
K_meanTable <- Kmean_generation(sumBioRepsMatrix, seed, k_clusters)
```
```{r showGeneratedKmeans, echo=FALSE}
knitr::kable(head(K_meanTable[, 1:4]),"pipe")
```

The most variable genes of each cluster are selected using the following function. A new object is created to save the most variable genes for each cluster, while the matrix that includes clusters for each gene is divided into individual cluster matrices. These matrices are then sorted by their variance. 
```{r exampleMostVariableGenes, echo=TRUE}
number_of_annotations_per_cluster <- 5
k_clusters <- 2
mostVarGeneslist <- most_variable_genes(K_meanTable, number_of_annotations_per_cluster, k_clusters)
print(mostVarGeneslist)
```
**To set the annotation for a heatmap the following function can be used:**

```{r exampleSetAnnotation, echo=TRUE}
number_of_annotations_per_cluster <- 5
annotation_for_heatmap <- set_annotation(sumBioRepsMatrix, number_of_annotations_per_cluster)
```

"performing_kMeans" is a function specifically written to perform k-means clustering outside of the general heatmap function. This allows for a more reliable splitting of the heatmap by its assigned clusters.
```{r examplePerformingKmeansClustering, echo=TRUE}
k_clusters <- 2
split_heatmap_clusters <- performing_kMeans(sumBioRepsMatrix, k_clusters)
```

```{r}
colorPalette <- "RdBu"
color_setting(colorPalette)
```

**Finally, a heatmap with clusters can be generated as in this example:**

```{r}
seed <- 1
title <- "Heatmap Cortex Iba1 positive"
fontsize_columnNames <-6
fontsize_rowNames <-4
title_heatmap_legend <- "Expression"
WidthNum <- 4.5
HeightNum <- 3
UnitSize <- "cm"
colorPalette <- "RdBu"
hm <- print_heatmap(seed, sumBioRepsMatrix, title, split_heatmap_clusters, annotation_for_heatmap, fontsize_columnNames, fontsize_rowNames, title_heatmap_legend, WidthNum, HeightNum, UnitSize, colorPalette)
```
With "function_complexHeatmap_var", a heatmap can be created with automatic annotation of the x most variable genes from each cluster. As input, a matrix with the scaled counts of the most variable genes from a dataset can be used. The function is able to summarize biological replicates.
```{r exampleFunctionComplexHeatmapVar, echo=TRUE}
title <- "Heatmap Cortex Iba1 positive"
fontsize_columnNames <-6
fontsize_rowNames <-4
title_heatmap_legend <- "Expression"
WidthNum <- 4.5
HeightNum <- 3
UnitSize <- "cm"
colorPalette <- "RdBu"

hm <- function_complexHeatmap_var(scaled_counts, probes, number_of_annotations_per_cluster, k_clusters, seed, title, fontsize_columnNames, fontsize_rowNames, title_heatmap_legend, WidthNum, HeightNum, UnitSize, colorPalette)
```

**A heatmap can further be generated with annotation of specific genes, as in this example:**

```{r exampleHeatmapSpecificAnnotation}
probes <- list("P0_cortex_Iba1_pos", "P5_cortex_Iba1_pos", "P15_cortex_Iba1_pos")
sumBioRepsMatrix <- summarise_bio_replicates(scaled_counts, probes)

seed <- 1
k_clusters <- 2
K_meanTable <- Kmean_generation(sumBioRepsMatrix, seed, k_clusters)

anno_specific_genes <- list("Stmn1", "Sox4", "Cldn5", "Mapt", "Camk2a", "Gapdh", "Plp1", "Apoe")
annotation_for_heatmap <- set_annotation(sumBioRepsMatrix, anno_specific_genes)
split_heatmap_clusters <- performing_kMeans(sumBioRepsMatrix, k_clusters)

title <- "Heatmap with specific annotation"
k_clusters <- 2
fontsize_columnNames <-6
fontsize_rowNames <-4
title_heatmap_legend <- "Expression"
WidthNum <- 4.5
HeightNum <- 3
UnitSize <- "cm"
colorPalette <- "RdBu"
hm <- print_heatmap(seed, sumBioRepsMatrix, title, split_heatmap_clusters, annotation_for_heatmap, fontsize_columnNames, fontsize_rowNames, title_heatmap_legend, WidthNum, HeightNum, UnitSize, colorPalette)

```

### Differential Gene Expression Analysis Using Nanostring Data

**Functions to analyze Nanostring GeoMx DSP data:**
```{r loadingPackagesNanostringData, warning=FALSE, include=FALSE, results='hide'}
library(NanoStringNCTools)
library(GeomxTools)
library("GeoMxWorkflows")
library(EnvStats)
library(ggiraph)
library(ggsankey)


if(packageVersion("GeomxTools") < "2.1" & 
   packageVersion("GeoMxWorkflows") >= "1.0.1"){
  stop("GeomxTools and Workflow versions do not match. Please use the same version. 
    This workflow is meant to be used with most current version of packages. 
    If you are using an older version of Bioconductor please reinstall GeoMxWorkflows and use vignette(GeoMxWorkflows) instead")
}

if(packageVersion("GeomxTools") > "2.1" & 
   packageVersion("GeoMxWorkflows") <= "1.0.1"){
  stop("GeomxTools and Workflow versions do not match. 
         Please use the same version, see install instructions above.")
  
  # to remove current package version
  # remove.packages("GeomxTools")
  # remove.packages("GeoMxWorkflows")
  # see install instructions above 
}

```

```{r readInNanostringRawFiles, echo=FALSE, results='hide', warning=FALSE}
datadir <- file.path("C:/Users/llancell/Documents/Paper")

DCCFiles <- dir(file.path(datadir, "dcc"), pattern = ".dcc$",
                full.names = TRUE, recursive = TRUE)

#has to be in file directory

PKCFiles <- file.path(datadir, "Mm_R_NGS_WTA_v1.0 (1).pkc")
PKCData <- dir(file.path(datadir, "pkc"), pattern = ".pkc$",
               full.names = TRUE, recursive = TRUE)

# has to be in file directory
SampleAnnotationFile <- file.path(datadir, "AllAnnotationNanostring1.xlsx")
```
The raw Nanostring GeoMx DSP files are read in and then united into an instance of class "NanostringGeoMxSet".
```{r ReadData, echo=TRUE, results='hide', warning=FALSE}
rawDataObject <- suppressWarnings(readNanoStringGeoMxSet(dccFiles = DCCFiles,
                                          pkcFiles = PKCFiles,
                                          phenoDataFile = SampleAnnotationFile,
                                          phenoDataSheet = "Template",
                                          phenoDataDccColName = "Sample_ID",
                                          protocolDataColNames = c("aoi","roi"),
                                          configFile = NULL,
                                          analyte = "RNA",
                                          phenoDataColPrefix = "",
                                          experimentDataColNames = NULL))


```
Anon Analyzing GeoMx-NGS RNA Expression Data with GeomxTools. Available from: 
http://bioconductor.riken.jp/packages/3.15/workflows/vignettes/GeoMxWorkflows/
inst/doc/GeomxTools_RNA-NGS_Analysis.html (September 21, 2023b). 
Anon Developer Introduction to the NanoStringGeoMxSet. Available from: 
https://www.bioconductor.org/packages/devel/bioc/vignettes/GeomxTools/inst/doc
/Developer_Introduction_to_the_NanoStringGeoMxSet.html (September 21, 2023c).

To preprocess the data, the logarithm of the count matrix is computed and added to the demoData object. Furthermore, the data is summarized by splitting it by a chosen column and calculating the mean.
```{r examplePreProcess, echo=TRUE, results='hide'}
PrePro_rawDataObject <- add_demoElem(rawDataObject)

class(PrePro_rawDataObject)
#loop over the features(1) or samples(2) of the assayData element and get the mean
assayDataApply(PrePro_rawDataObject, MARGIN=1, FUN=mean, elt="demoElem")[1:5]

#split the data by group column with feature, pheno or protrocol data then get the mean
VGroup <- "aoi"
elt <- "demoElem"
PrePro_rawDataObject <- split_data_by_column(PrePro_rawDataObject, VGroup, elt)

# Quality Control
QCPassed <- aExprsDataQC(PrePro_rawDataObject, "QCFlags")

df_Exp <- genRawReadCountTable(PrePro_rawDataObject)
```
```{r showExpressionMatrix, echo=FALSE, warning=FALSE}
knitr::kable(head(df_Exp[, 1:9]),"pipe")
``` 
Anon Analyzing GeoMx-NGS RNA Expression Data with GeomxTools. Available from: 
http://bioconductor.riken.jp/packages/3.15/workflows/vignettes/GeoMxWorkflows/
inst/doc/GeomxTools_RNA-NGS_Analysis.html (September 21, 2023b). 
Anon Developer Introduction to the NanoStringGeoMxSet. Available from: 
https://www.bioconductor.org/packages/devel/bioc/vignettes/GeomxTools/inst/doc
/Developer_Introduction_to_the_NanoStringGeoMxSet.html (September 21, 2023c).

Next, an usable matrix of column data is extracted from the sample data:
```{r ColdataDEG, echo=TRUE, results='hide', warning=FALSE}
annotation_matrix <- sData(rawDataObject)

annotation_matrix$Samplename <- paste(annotation_matrix$age_region, annotation_matrix$segment, annotation_matrix$sample, sep="_")
class(annotation_matrix)
library(tibble)

annotation_matrix <- tibble::rownames_to_column(annotation_matrix, "Samplenummern")

rownames(annotation_matrix)

annotation_matrix1 <- data.frame(annotation_matrix$Samplenummern, annotation_matrix$Samplename)

coldata_2 <- data.frame(annotation_matrix$Samplename, annotation_matrix$age, annotation_matrix$region, annotation_matrix$segment, annotation_matrix$sample)

rownames(coldata_2) <- coldata_2[,1]
# Renaming the columns
names(coldata_2)[names(coldata_2) == "annotation_matrix.Samplename"] <- "Samplename"
names(coldata_2)[names(coldata_2) == "annotation_matrix.age"] <- "age"
names(coldata_2)[names(coldata_2) == "annotation_matrix.region"] <- "region"
names(coldata_2)[names(coldata_2) == "annotation_matrix.segment"] <- "segment"
names(coldata_2)[names(coldata_2) == "annotation_matrix.sample"] <- "sample"

# Combines sample describing columns into one
coldata_2$comp <- paste(coldata_2$age, coldata_2$region, coldata_2$segment, sep = "_")

coldata_2 <- coldata_2[,-1]
coldata_2 <- as.matrix(coldata_2)

# Creates a new dataframe wich only includes the necessary information
coldata_df <- as.data.frame(coldata_2)
class(coldata_df)

```
The columns in the counts table can now be replaced with the corresponding sample names from the table containing the column data. To do this, the following code can be used:
```{r exampleExchangeColumnNamesWMatchingSampleNames, echo=TRUE, results='hide'}
list_columnNames <- list(colnames(df_Exp)) # list of column names in counts table
print(list_columnNames)
annotationMatrix <- annotation_matrix1 # creates copy of column data 
rownames(annotationMatrix) <- annotationMatrix[,2] # makes values in second colum into rownames
x <- match(rownames(annotationMatrix), colnames(df_Exp))
print(x)
copy_df_Expr <- df_Exp # creates copy of counts table

for (i in list_columnNames) {
  matchingRow <- which(annotationMatrix$annotation_matrix.Samplenummern == i)
  matchingColumn <- match(i, names(copy_df_Expr))
  print(matchingColumn)
  print(matchingRow)
  colnames(copy_df_Expr)[matchingColumn] <- rownames(annotationMatrix)[matchingRow]
  print(rownames(annotationMatrix)[matchingRow])
  }
```
```{r exampleColumnNameExchange, echo=FALSE}
knitr::kable(head(copy_df_Expr[, 1:9]),"pipe")
```


**Differential gene expression analysis**

The process of performing a differential gene expression analysis starts by checking whether all the columns of the raw data are present in the metadata and if the column names appear in the same order in both the raw data and metadata.


```{r EqualOrder, echo=TRUE, warning=FALSE}
all(colnames(copy_df_Expr) %in% rownames(coldata_2)) #check if all column names of data are in rownames of metadata #check if TRUE
class(coldata_2)

all(colnames(copy_df_Expr) == rownames(coldata_2)) #check if the order of the data column names == order of metadata rownames

```
Love M, Ahlmann-Eltze C, Forbes K, Anders S, Huber W, FP7 RE, NHGRI N, CZI (2023) DESeq2: 
Differential gene expression analysis based on the negative binomial distribution. 
https://doi.org/10.18129/B9.bioc.DESeq2

To perform the differential gene expression analysis, limma voom is utilized. First, a matrix is created in which each coefficient represents the average of the samples from one group. Afterwards, the contrasts are set to direct testing between chosen groups in the analysis. To scale the raw library sizes, the normalization factors are calculated. The count data is then transformed to log2-counts per million, and the mean-variance relationship is estimated. By fitting a linear model to each gene, both fold changes and standard errors are estimated. The fitted model objects are then reorientated from the coefficients of the design matrix to any set of contrasts of the original coefficients. The standard errors are smoothed by empirical Bayes. Finally, the results of the differential gene expression analysis are classified as up, down, or not significantly differently expressed.
```{r PerformingDEGLimmaVoom,echo=TRUE, include=TRUE, error=TRUE, warning=FALSE}
library(edgeR)
library(dplyr)
library(tidyr)
library(tidyverse)

coldata2 <- as.data.frame((coldata_2))
comp <- paste0(coldata2$age, coldata2$region, coldata2$segment)
design <- model.matrix(~0 + comp)


contrasts <- makeContrasts(P0_Iba1_neg_cortex_vs_cc = compP0cortexIba1_neg - compP0ccIba1_neg,
                           P0_Iba1_neg_cortex_vs_hippo = compP0cortexIba1_neg - compP0hippoIba1_neg,
                           P0_Iba1_neg_cc_vs_hippo = compP0ccIba1_neg - compP0hippoIba1_neg,
                           levels = design)

y = DGEList(counts = copy_df_Expr)
y = calcNormFactors(y)
v = voom(y, design)

fit <- lmFit(v, design) 
fit2 <- contrasts.fit(fit, contrasts) # contrasts.fit must be run before eBayes
fit2 <- eBayes(fit2)

results_all_DEG <- decideTests(fit2)
summary(results_all_DEG)

result1 = topTable(fit2, coef= "P0_Iba1_neg_cc_vs_hippo", number = Inf, adjust.method = "fdr") %>%
  as.data.frame() # differnetially expressed genes are obtained by topTreat() function
# filter results to get significantly differentially expressed genes




topUp <- result1[which(result1$logFC > 0),] [1:100,] # up reg top 100
topDown <- result1[which(result1$logFC < 0),] [1:100,] # down reg top 100

class(topUp)

NamesUpReg <- row.names(topUp)
NamesDownReg <- row.names(topDown)
print(NamesDownReg)

result1 <- result1 %>%
  dplyr::mutate(isSignificant = case_when(
    adj.P.Val < 0.05& abs(logFC) >1 ~TRUE,
    TRUE ~ FALSE # If condictions in the line above are not met, gene is not DE.
    ))

sigDEresults <- result1 %>%
  dplyr::filter(isSignificant == TRUE)

```
inesdesantiago (2020) 10 Tips & Tricks for complex model.matrix designs in DGE analysis. 
Sequencing QC and data analysis blog. Available from: 
Bibliography
96
https://seqqc.wordpress.com/2020/11/28/10-tips-tricks-for-complex-model-matrixdesigns-in-dge-analysis/ (September 21, 2023).
Anon Developer Introduction to the NanoStringGeoMxSet. Available from: 
https://www.bioconductor.org/packages/devel/bioc/vignettes/GeomxTools/inst/doc
/Developer_Introduction_to_the_NanoStringGeoMxSet.html (September 21, 2023c). 


**Visualisation with a Venn Diagramm:**

The results from the differential gene expression analysis are visualized using a Venn diagramm to represent the up- and down- regulated genes between groups. This is set up in the following way:
```{r VennDiagramm, echo=TRUE, error=TRUE, warning=FALSE}
library(VennDiagram)
venn.plot <- vennDiagram(results_all_DEG,
              imagetype = "tiff",
              include=c("up", "down"), mar=rep(1,4), cex=c(1.5,1,0.7), lwd=1,
              counts.col=c("red", "blue"),
              circle.col = c("red", "blue", "green3"))
```
Chen, H., Boutros, P.C. VennDiagram: a package for the generation of highly-customizable Venn and Euler diagrams in R. BMC Bioinformatics 12, 35 (2011). https://doi.org/10.1186/1471-2105-12-35

