---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# DgeaHeatmap

<!-- badges: start -->
<!-- badges: end -->

The goal of DgeaHeatmap is to enable R users to generate heatmaps more easily and to help with preprocessing read counts. Furthermore, the package is aimed to simplify the extraction of raw read counts from .dcc and .pkc files generated through Nanostring GeoMx DSP.

## Installation

You can install the development version of DgeaHeatmap from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("leolanci/Dgea_Heatmap_Package")
```

## Usage

This is a basic example which shows you how to solve a common problem:
```{r input normalizedReadCountsExample, echo=FALSE, results='hide'}
library(testthat)
input_data <- read.csv(test_path("1_Counts_All_Regions_All_Ages_Pos_Neg.csv"))
del_layer6 <- grep("layer", colnames(input_data))
input_data <- input_data[, -del_layer6]
```


```{r exampleBuildMatrix, echo=TRUE, results='hide'}
library(DgeaHeatmap)
x <- 1
matrixCounts <- build_matrix(input_data, x)

```
```{r showTable, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, out.height="60%", out.width="60%"}
knitr::kable(head(matrixCounts[, 1:4]),"pipe")
```
```{r exampleIndividualMatrix, echo=TRUE}
parameter1 = "cortex"
parameter2 = "pos"
factors_for_individual_matrix = list(parameter1, parameter2)
indiMatrix <- individual_matrix(factors_for_individual_matrix, matrixCounts)
```
```{r showIndividualMatrix, echo=FALSE}
knitr::kable(head(indiMatrix[, 1:9]),"pipe")
```

Filtering the matrix for only an x amount of most variable expressed genes:
```{r exampleFilteringForTopExprGenes, echo=TRUE}
top_number_of_genes <- 500
varGenesMatrix <- filtering_for_top_exprGenes(indiMatrix, top_number_of_genes)
print(nrow(varGenesMatrix))
```

Using Z-score scaling to scale the values of the matrix:
```{r exampleZScoreScaling, echo=TRUE, results='hide'}
scaled_counts <- scale_counts(varGenesMatrix)
```

How to visualize the data distribution of the scaled counts:
```{r exampleShowDataDistribution, echo=TRUE}
show_data_distribution(scaled_counts)
```
Generating an elbow plot to choose k for k-mean generation:
```{r exampleElbowPlot, echo=TRUE}
seed <- 1 # setting a seed for a reproducible outcome
elbow_plot(seed, scaled_counts)
```
As desired, the samples can further be summarized as biological replicates to help generate more clearly arranged heatmaps and giving a better overview.
```{r exampleSummariseBioReplicates, echo=TRUE, results='hide'}
probes <- list("P0_cortex_Iba1_pos", "P5_cortex_Iba1_pos", "P15_cortex_Iba1_pos")
sumBioRepsMatrix <- summarise_bio_replicates(scaled_counts, probes)
```
```{r showSummedBioReplicates, echo=FALSE}
knitr::kable(head(sumBioRepsMatrix[, 1:3]),"pipe")
```
Generating K-means for clustering in the heatmap:
```{r exampleKmeanGeneration, echo=TRUE}
seed <- 1 #set a seed for reproducibility, but try different seeds first
k_clusters <- 2
K_meanTable <- Kmean_generation(sumBioRepsMatrix, seed, k_clusters)
```
```{r showGeneratedKmeans, echo=FALSE}
knitr::kable(head(K_meanTable[, 1:4]),"pipe")
```

The most variable genes of each cluster are selected through the following function. A new object is created with the purpose of saving the most variable genes for each cluster, while the matrix with included cluster for each gene is divided into individual cluster matrices. These matrices are each sorted by their variance. 
```{r exampleMostVariableGenes, echo=TRUE}
number_of_annotations_per_cluster <- 5
k_clusters <- 2
mostVarGeneslist <- most_variable_genes(K_meanTable, number_of_annotations_per_cluster, k_clusters)
print(mostVarGeneslist)
```
To set the annotation for a heatmap the following function can be used:
```{r exampleSetAnnotation, echo=TRUE}
number_of_annotations_per_cluster <- 5
annotation_for_heatmap <- set_annotation(sumBioRepsMatrix, number_of_annotations_per_cluster)
```

"performing_kMeans" is a function specifically written to perform k-mean clustering outside of the general heatmap function. This allows a more reliable splitting of the heatmap by their assigned clusters.
```{r examplePerformingKmeansClustering, echo=TRUE}
k_clusters <- 2
split_heatmap_clusters <- performing_kMeans(sumBioRepsMatrix, k_clusters)
```

```{r}
colorPalette <- "RdBu"
color_setting(colorPalette)
```

Finally, a heatmap with clusters can be generated as in this example:
```{r}
seed <- 1
title <- "Heatmap Cortex Iba1 positive"
fontsize_columnNames <-6
fontsize_rowNames <-4
title_heatmap_legend <- "Expression"
WidthNum <- 4.5
HeightNum <- 3
UnitSize <- "cm"
colorPalette <- "RdBu"
print_heatmap(seed, sumBioRepsMatrix, title, split_heatmap_clusters, annotation_for_heatmap, fontsize_columnNames, fontsize_rowNames, title_heatmap_legend, WidthNum, HeightNum, UnitSize, colorPalette)
```
With "function_complexHeatmap_var" a heatmap can be created with automatic annotation of x most variable genes from each cluster. As input a matrix with the scaled counts of the most varible genes from a dataset can be used. The function is able to summarize biological replicates.
```{r exampleFunctionComplexHeatmapVar, echo=TRUE}
title <- "Heatmap Cortex Iba1 positive"
fontsize_columnNames <-6
fontsize_rowNames <-4
title_heatmap_legend <- "Expression"
WidthNum <- 4.5
HeightNum <- 3
UnitSize <- "cm"
colorPalette <- "RdBu"

hm <- function_complexHeatmap_var(scaled_counts, probes, number_of_annotations_per_cluster, k_clusters, seed, title, fontsize_columnNames, fontsize_rowNames, title_heatmap_legend, WidthNum, HeightNum, UnitSize, colorPalette)
```
A heatmap can further be generated with annotation of specific genes, as in this example:
```{r exampleHeatmapSpecificAnnotation}
probes <- list("P0_cortex_Iba1_pos", "P5_cortex_Iba1_pos", "P15_cortex_Iba1_pos")
sumBioRepsMatrix <- summarise_bio_replicates(scaled_counts, probes)

seed <- 1
k_clusters <- 2
K_meanTable <- Kmean_generation(sumBioRepsMatrix, seed, k_clusters)

anno_specific_genes <- list("Stmn1", "Sox4", "Cldn5", "Mapt", "Camk2a", "Gapdh", "Plp1", "Apoe")
annotation_for_heatmap <- set_annotation(sumBioRepsMatrix, anno_specific_genes)
split_heatmap_clusters <- performing_kMeans(sumBioRepsMatrix, k_clusters)

title <- "Heatmap with specific annotation"
k_clusters <- 2
fontsize_columnNames <-6
fontsize_rowNames <-4
title_heatmap_legend <- "Expression"
WidthNum <- 4.5
HeightNum <- 3
UnitSize <- "cm"
colorPalette <- "RdBu"
print_heatmap(seed, sumBioRepsMatrix, title, split_heatmap_clusters, annotation_for_heatmap, fontsize_columnNames, fontsize_rowNames, title_heatmap_legend, WidthNum, HeightNum, UnitSize, colorPalette)

```
<font size="5"> Functions to analyze Nanostring GeoMx DSP data</font>
```{r loadingPackagesNanostringData, echo=FALSE, results='hide'}
library(NanoStringNCTools)
library(GeomxTools)
library("GeoMxWorkflows")
library(EnvStats)
library(ggiraph)
library(ggsankey)


if(packageVersion("GeomxTools") < "2.1" & 
   packageVersion("GeoMxWorkflows") >= "1.0.1"){
  stop("GeomxTools and Workflow versions do not match. Please use the same version. 
    This workflow is meant to be used with most current version of packages. 
    If you are using an older version of Bioconductor please reinstall GeoMxWorkflows and use vignette(GeoMxWorkflows) instead")
}

if(packageVersion("GeomxTools") > "2.1" & 
   packageVersion("GeoMxWorkflows") <= "1.0.1"){
  stop("GeomxTools and Workflow versions do not match. 
         Please use the same version, see install instructions above.")
  
  # to remove current package version
  # remove.packages("GeomxTools")
  # remove.packages("GeoMxWorkflows")
  # see install instructions above 
}

```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
